### 2017-09-26 paper/article_0626.tex
adds a bunch of writing changes

\section{Introduction}
Modeling for disease surveillance data to explore patterns of infectious diseases has had a long history in public health. Infectious diseases can cause high economic and medical costs due to morbidity and mortality. In recent decades, the annual number of global deaths caused by infections has leveled off at approximate 15 million and may remain at this level for the next three decades \cite{DyeC, WHOM}. In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the course of disease transmission. Therefore, quantifying the dynamics of infection is the key to help us better understand zoonoses, which also provides valuable information to professionals who are making decisions on public health policy for disease control.

Over the last few decades there have been many models developed for the analysis of zoonotic diseases. Epidemiological models make use of proportions of cases and geographic data to capture the main characteristics of the spread system. For example, the Susceptible-Infectious-Recovered model \cite{Kerm} classifies populations into a susceptible, infective and recovered status enabling an estimation of the number of individuals who are likely to be infected. Also, models with a combination of spatial and temporal data enable the prediction of the time frame and areas a disease or an outbreak is likely to occur \cite{HeldL, Hoehl, Simo}. However, a framework of a zoonotic transmission passed from animals to humans can have a complexity that involves various exposures contaminated via different pathways. For instance, infected wild birds may contaminate environmental water and cause the spread to water users, either humans or non-humans \cite{Wagen}. This implies that tracing back the source of infection is crucial in enriching the resolution of the network, thereby increasing the ability to implement risk management and intervention \cite{Wilso, Morel}. In terms of statistical analysis, it also means that modeling for zoonoses requires an advanced approach with the focus changed from epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}. Hence, some source attribution models have been proposed to estimate the number of cases attributable to different sources by utilizing the epidemiological information and the association with genotypes found in humans and sources \cite{vanP, Hald, MullA}.

@@

The genetic information used in the integrated models is typically the genotypes that capture the dynamic changes on DNA sequences occurring in the transmitted pathogens \cite{Cotta}. In order to incorporate the genetic information in the epidemiological data, some molecular tools may need to be deployed in the analysis. For example, a multilocus sequence typing (MLST) \cite{Dingl, Coll} is developed and used particularly to identify the species of the pathogen \textit{Campylobacter}, which causes campylobacteriosis, a worldwide gastrointestinal disease in humans. This method uses the location of genes (called loci) and the variants of genes (called alleles, caused by mutation and recombination) observed from fragmented DNA sequences typed from isolates, which are subjects taken from groups of bacterial colonies produced by a small piece of sample in culture plates. In the process of type identification, MLST determines seven specific housekeeping genes (or loci), representing the main functions of the whole genome sequence, to characterize the species of \textit{Campylobacter} \cite{Urwin}. These housekeeping genes are \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit). Once the seven loci are targeted on sequences, MLST produces an allelic profile consisting of seven positive integers addressing the unique genetic sequence at the respective locus, and it also assigns a unique positive integer as a sequence type (ST) to the associated profile for designation. An illustrative example of a part of MLST data is presented in Table~(\ref{tab0}). It shows that the genotypes ST-45 and ST-474 have different allelic combinations across the seven loci. The different allelic profiles enable comparison of gene similarities or dissimilarities so that an association between sources and infected cases can be made.

The epidemiological and molecular-based models provide greater insight into source attribution. It improves knowledge about infection attributed to sources from the time scale and the location point of view if spatial-temporal covariates are considered in the model. However, few authors in the literature put an emphasis on validations to these integrated models, in which the genetic effects are difficult to assess analytically due to increased model complexity. We are therefore motivated to develop a simpler model so that we can evaluate the integrated source attribution model and examine its rationality for depicting the actual attribution.

In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration. Human campylobacteriosis is caused by \textit{Campylobacter}, of which \textit{C. jejuni} and \textit{C. coli} are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}. The common symptoms of infection are diarrhea, abdominal pain and fever. However, a severe complication named Guillain Barre syndrome might be developed, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}. The pathogen can be spread between animals, or from animals and wild birds to humans. The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces. In the study, the applied model \cite{Wilso} particularly considers genetic evolution in the pathogen that gives an estimation of the proportion of cases that are more likely to get infected from sources when some genotypes are observed from infected individuals. To uncover the genetic effects contributed behind the model, a simple model is particularly developed to simulate comparative information as the integrated model provides. Based on the information produced by the two models we estimate and compare the attribution probability for each source by fitting a general model, from which more covariates can be incorporated, to explore the variation of the attribution probability from a fresh perspective. Last, we apply the approach to the case study to explore the differences in source attribution between the integrated model and the simple model in relation to the geographical areas where the human cases resided.

\section{Models for source attribution with/without information of evolutionary change}
\subsection{Notation and assumptions}
The ultimate goal of the model development is to estimate the source attribution probability, which can quantify the attribution to each source with probability given a genotype is identified in an individual. The main idea behind the establishment has two parts: the sampling distribution of genotypes associated with sources, and the proportion of human cases for each source. Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.

>>>>

\section{Introduction}
Modeling for disease surveillance data to explore patterns of infectious diseases has had a long history in public health. Infectious diseases can cause high economic and medical costs due to morbidity and mortality. In recent decades, the annual number of global deaths caused by infections has leveled off at approximate 15 million and may remain at this level for the next three decades \cite{DyeC, WHOM}. In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the course of disease transmission. For zoonotic diseases, where transmission to humans may be via food, water, direct contact with animals, or through environmental contamination, knowledge of the potential sources and pathways of infection is key in improving understanding of the burden of disease. Quantifying the burden of illness to each pathway and source, provides valuable information to health professionals making decisions on public health policy for disease control.

Over the last few decades there have been many models developed for the analysis of zoonotic diseases. Epidemiological models make use of the incidence of cases and geographic data to capture the main characteristics of disease spread. For example, the Susceptible-Infectious-Recovered model \cite{Kerm} classifies populations into a susceptible, infective and recovered status enabling an estimation of the number of individuals who are likely to be infected. Also, models with a combination of spatial and temporal data enable the prediction of the time frame and areas where a disease outbreak is likely to occur \cite{HeldL, Hoehl, Simo}. However zoonotic transmission from animals to humans may be complex, involving many sources and exposures linked by differing pathways. For instance, infected wild birds may contaminate environmental water and cause the spread to water users, either humans or non-humans \cite{Wagen}. Tracing the source of infection is thus crucial to increasing the ability to implement risk management and intervention \cite{Wilso, Morel}. In terms of statistical analysis, it means that modelling zoonoses requires an advanced approach with the focus changed from just epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}. Hence, some source attribution models have been proposed to estimate the number of cases attributable to different sources by utilizing epidemiological information and the association with genotypes found in humans and sources \cite{vanP, Hald, MullA}.

@@

The genetic information used in such integrated models is typically at the level of genotyping that groups closely related organisms together to capture broader changes in DNA occurring in the transmitted pathogens \cite{Cotta}. Typically molecular typing methods, such as multilocus sequence typing (MLST) \cite{Dingl, Coll} might be used, which utilises nucleotide sequences of internal fragments of a small set of housekeeping genes. Such sequences have sufficient variation to distinguish differing pathogen lineages, while being relatively stable within lineages. Each unique nucleotide sequence (allele) at each housekeeping gene (loci) is assigned a number, and the set of numbers across all loci (the allelic profile) is then taken as the genotype, with is usually also assigned a number.

For the pathogen \textit{Campylobacter}, which causes campylobacteriosis, a worldwide gastrointestinal disease in humans. For \textit{Campylobacter}, the MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit). An illustrative example of a part of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab0}). It shows that the genotypes ST-45 and ST-474 have different allelic combinations across the seven loci. The different allelic profiles enable comparison of gene similarities or dissimilarities so that an association between sources and infected cases can be made.

The first step for attribution models that utilise genetic information is building the sampling distribution of genotypes among each putative source. This may range from using the proportion of each observed genotype (CITATION: TINA HALD) on each source through to utilising allelic profile information to derive mutation and recombination rates within each source, and migration rates between each source (CITATION: ISLAND MODEL). A key question is whether more complex genetic models yield superior attribution results or whether a significantly simpler model may suffice, however few authors in the literature have addressed this point. This becomes more important as model complexity extends to include epidemiological covariates. We are therefore motivated to develop a simple genetic model in order to assess the additional information that the more complex models provide.

In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration. Human campylobacteriosis is caused by \textit{Campylobacter}, of which \textit{C. jejuni} and \textit{C. coli} are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}. The common symptoms of infection are diarrhea, abdominal pain and fever. However, a severe complication named Guillain Barre syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}. The pathogen can be spread between animals, or from animals and wild birds to humans. The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces. In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilises only the prevalence of each type to derive the sampling distribution. We then extend both models to incorporate covariates, exploring the effect of human case rurality on attribution results, modelling attribution by rurality using a linear trend or using separate categories and performing model comparison.

\section{Models for source attribution with and without information of evolutionary change}
\subsection{Notation and assumptions}
The ultimate goal of the model development is to estimate the source attribution probability, which can quantify the attribution to each source with probability given a genotype is identified in an individual. The main idea behind the establishment has two parts: the sampling distribution of genotypes associated with sources, and the proportion of human cases for each source. Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.
